Variety in the Implementation of Nonlinear Least Squares Program Codes

### 2022-06-13 RefactoringNLS/RefactoringNLS.Rmd
Updates to two articles

>>>>

??fix next bits
<!-- - The different dialects of modeling expressions as detailed above should be rationalized. This -->
<!--   would, however, be a painful and mainly political activity. -->

<!-- - The tools for analytic derivatives that exist in parts of `nls()` for self-starting models and -->
<!--   in `nlsr` for problems specified as models that use expressions for which derivatives exist  -->
<!--   should be more generally available, but the task is one of tedium and limited reward to the -->
<!--   persons who do it, while benefiting the R community substantially. -->

<!-- - Similarly, default methods need stabilized solvers e.g., Marquardt methods, though to date -->
<!--   a mechanism to incorporate such modifications in `nls()` has eluded us. -->

<!-- - Bounds constraints are sufficiently important that they should be part of the default -->
<!--   methods -->
<!--   (rather than algorithm = "port" for `nls()`). `minpack.lm` should handle them correctly. -->

### 2022-06-13 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
Updates to two articles

# Test problems

It is worth including some sample test problems in this article to illustrate
the discussion.

?? here or later

```{r settest15, fig.width=5}
# set parameters
set.seed(123456)
a <- 1
b <- 2
c <- 3
np <- 15
tt <-1:np
yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
# plot.new()
plot(tt, yy)
#plot(tt, yy, type='l')
set.seed(123456)
ev <- runif(np)
ev <- ev - mean(ev)
y1 <- yy + ev
points(tt,y1,type='p', col="green")
y2 <- yy + 5*ev
points(tt,y2,type='p', col="blue")
y3 <- yy + 10*ev
lg3d15 <- data.frame(tt, yy, y1, y2, y3)
 points(tt,y3,type='p', col="red")
library(nlsr)
sol0 <- nlxb(yy ~ a0/(1+b0*exp(-c0*tt)), data=lg3d15, start=list(a0=1, b0=1, c0=1))
print(sol0)
sol1 <- nlxb(y1 ~ a1/(1+b1*exp(-c1*tt)), data=lg3d15, start=list(a1=1, b1=1, c1=1))
print(sol1)
sol2 <- nlxb(y2 ~ a2/(2+b2*exp(-c2*tt)), data=lg3d15, start=list(a2=1, b2=1, c2=1))
print(sol2)
sol3 <- nlxb(y3 ~ a3/(3+b3*exp(-c3*tt)), data=lg3d15, start=list(a3=1, b3=1, c3=1))
print(sol3)
```

The following is a larger dataset version of this test.

```{r settest150, eval=FALSE}
np <- 150
tt <- (1:np)/10
yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
set.seed(123456)
ev <- runif(np)
ev <- ev - mean(ev)
y1 <- yy + ev
y2 <- yy + 5*ev
y3 <- yy + 10*ev
lg3d150 <- data.frame(tt, yy, y1, y2, y3)
np <- 1500
tt <- (1:np)/100
yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
set.seed(123456)
ev <- runif(np)
ev <- ev - mean(ev)
y1 <- yy + ev
y2 <- yy + 5*ev
y3 <- yy + 10*ev
lg3d1500 <- data.frame(tt, yy, y1, y2, y3)
f0 <- yy ~ a0/(1+b0*exp(-c0*tt))
f1 <- y1 ~ a1/(1+b1*exp(-c1*tt))
f2 <- y2 ~ a2/(2+b2*exp(-c2*tt))
f3 <- y3 ~ a3/(3+b3*exp(-c3*tt))
```

>>>>

# Summary discussion

This article has attempted to provide an overview of how software for nonlinear least
squares computations may be structured and implemented, taking note that users may wish
to solve modeling problems rather than minimize a sum of squared functions. We have seen
that different implementation goals and structural decisions greatly influence the
nature, capability, performance and features of different codes. Moreover, while the
treatment here has taken a panoramic view, we have been interested primarily in what
is availabel to R users.

Having reviewed the tools available, we note that there are some serious concerns. Of these,
we will list those we consider to be the most important:

- tools such as the `nls()` function in R has so many legacy uses that it is likely
  impossible to deprecate it. However, it has also (as we detail in ??RefactoringNLS??)
  non-maintainable, but it seems unlikely that any
  further development, nor even repair of current deficiencies, will be forthcoming.
  This underlines the importance of designing software for its long-term maintenance,
  and this is often inimical to using currently novel programming tools.

- The rich variety of ways of structuring and solving the Gauss-Newton equations, stabilized
  or not, allows for codes to be implemented to take advantage of special situations. 
  However, for most use-cases, an approach using reliable matrix decompositions such
  as modern QR techniques should suffice. These allow Marquardt stabilizations to be
  applied relatively easily in most, but not all, situations.

- The needs of different user leads to difficult choices and implementation methods for
  handling how problems are interfaced to the computational infrastructure and results
  are returned and reported. These need to be well-documented, paying attention to 
  the different ways in which users understand their problems.
